---
title: Spatial modeling with INLA
subtitle: A short tutorial
author: B. K. M. Case
date: April 11, 2022
output: rmarkdown::github_document
---

Here we provide a short tutorial on fitting the types of spatial models used throughout the paper using the R-INLA package. 
INLA (Integrated Nested Laplacian Approximation) is an alternative to MCMC for fitting a wide variety of Bayesian models, 
and has become a popular option for geospatial models because of its speed and relative ease of use 
(see [Bakka et al. (2018)](https://wires.onlinelibrary.wiley.com/doi/pdf/10.1002/wics.1443) for a recent review).

### Load required packages and the data

```{r, include=FALSE}
knitr::opts_chunk$set(fig.path="readme-figs/")
```

```{r, message=FALSE, results=FALSE, warning=FALSE}
library(INLA)
library(tidyverse)
source("code/seq-sampling-helpers.R")


all <- read_csv("anon-survey-data.csv")
pat <- filter(all, village == "Paternito")
```

### Patterns of infestation

We are going to fit a spatial model to the infestation data in El Paternito. To begin exploring this data, we
can produce a plot of the binary infestation status of each home. Note that the latitude and longitude are not actual coordinates
here as they have been anonimized, but we can still treat them as x and y coordinates, respectively.

```{r}
ggplot(pat, aes(long, lat, col = as.factor(infestation))) +
    geom_point() +
    labs(x = NULL, y = NULL, col = 'Infestation status')
```

We find that infested houses tend to be in middle and upper parts of the village, and often can be found
 close to other infested houses. 

### Traditional logistic regression

Before incorporating spatial effects, let's fit a model using only fixed effects from the covariate information.  
For simplicity, let's only use the covariates `bed_hygiene`, `num_dogs`, and `dist_perim`.
To fit such a model in R-INLA, we use the `inla` function:

```{r}
fit_fx_only <- inla(infestation ~ bed_hygiene + num_dogs + dist_perim, data = pat, family = 'binomial')
summary(fit_fx_only)
```

You may notice the syntax is quite similar to base R's `lm` and `glm` functions: first we provide a formula with names corresponding 
to the data specified with `data = pat`, then we specify a binomial response to indicate our data is binary. Notice also that, just like base R, INLA 
will automatically convert discrete variables into dummy variables in the model matrix.

This is all it takes to perform fully Bayesian inference using INLA! Note that here we have used the default priors for the fixed effect coefficients. 

#### Summary statistics and marginal distributions

In addition to viewing some summary statistics of the posterior with the `summary` function, 
this information is contained as fields in the INLA fit object:

```{r}
fit_fx_only$summary.fixed$`0.5quant` # median of fixed effects coefficients
head(fit_fx_only$summary.linear.predictor) # "untransformed" risk of infestation
head(fit_fx_only$summary.fitted.values) # transformed risk (between 0 and 1)
```

Notice the distinction between `linear.predictor`, the logit transformed risk, and `fitted.values`, which is the actual risk of infestation.
Now we plot the expected risk of each home compared to its true infestation status:

```{r}
pat |>
    mutate(pred_risk = fit_fx_only$summary.fitted.values$mean) |>
    ggplot(aes(long, lat, col = as.factor(infestation), size = pred_risk)) +
    geom_point() +
    labs(x = NULL, y = NULL, col = 'Infestation status', size = "Pred. risk")
```

Marginal distributions for the coefficients can be accessed similarly, which can then be converted into density functions, sampled from, etc.

```{r}
marg <- fit_fx_only$marginals.fixed$dist_perim
qplot(marg[,1], marg[,2], xlab = "Distance from perimeter", ylab = "Density")
```

The posterior probability that `dist_perim` has a positive association with the response can be computed like this:

```{r}
1 - inla.pmarginal(0, marg)
```

A number of other functions for working with marginal distributions are available.

### Adding random effects in R-INLA

Adding random effects to your hierarchical model is easy and where the INLA approach really starts to shine. The following models are 
supported:

```{r}
names(inla.models()$latent)
```

Documentation and examples on each of these models can be found using `inla.doc("<model name>")`.

Random effects are included with the `f` function in the model formula. This function takes an index variable as its first argument,
which is used to assign observations to groups. For example, to include `village` as a random effect when using the data from all villages, simply 
use this variable as the index variable:

```{r}
base_f <- infestation ~ bed_hygiene + num_dogs + dist_perim
fit_vil <- inla(update(base_f, ~. + f(village, model = "iid")), data = all, family = "binomial")
summary(fit_vil)
```

There is now a new section in the model summary: the *hyperparameters*. In INLA these are defined as any unknown variables that the random
effects depend on. Notice the precision(inverse variance) of our random effect itself has an enormous variance. This is because the 
default prior for the precision is quite diffuse. While this is an appropriate weakly informative prior when using linear regression, this is far 
too diffuse for good shrinkage using logistic regression and our particular data set. Adjusting these priors is straight forward, but note they are 
defined on log scale with the transformation `theta = log(precision)`.

Similarly, let's adjust the priors on the fixed effect coefficients to have a more suitable variance for logistic regression.

```{r}
hyper_pri_config <- list(theta = list(prior = "loggamma", param = c(1, 0.01)))
fx_pri_config <- list(mean.intercept = 0, prec.intercept = 0.05, mean = 0, prec = 0.05)
fit_vil <- inla(
    update(base_f, ~. + f(village, model = "iid", hyper = hyper_pri_config)), 
    data = all, family = "binomial",
    control.fixed = fx_pri_config
)
summary(fit_vil)
```

This fit appears much better, as evident from the higher marginal likelihood and lower posterior variance on the precision.

### Spatial smoothing with `dmatern`

We now turn to adding spatially correlated random effects. INLA supports several models suitable for spatial smoothing on 
regular grids (e.g. `rw2`), continuous points (e.g. `spde2`), and discrete regions (e.g. `bym`). Here we use a model for continuous 
points since our data are individual points occuring at irregular intervals.

The most popular way to model continuous points in space with INLA is the SPDE (Stochastic Partial Differential Equation) approach. 
This formulation requires more setup, but is much faster than traditional formulations because it approximates the dense covariance matrix
between points with a sparse matrix. 

For brevity, and due to many helpful tutorials on using SPDE with INLA already being online (several are listed below), we instead focus on 
a different formulation available, the `dmatern` model. This model directly represents the covariance between points and is equivalent to using SPDE.

Here, the indexing variable `idx` for `f` will need to be a unique integer for each house, since each house has one observation. Additionally
we will pass a matrix where each row `i` contains the x y coordinates of the house indexed with `idx = i`.

```{r}
idx <- 1:nrow(pat) # indexes rows of location matrix
loc_mat <- as.matrix(select(pat, long, lat)) # matrix of x y coordinates for each point
  
fit_gp_dense <- inla(
    update(base_f, ~. + f(idx, model = "dmatern", locations = loc_mat)),
    data = mutate(pat, idx = idx),
    family = 'binomial',
    control.fixed = fx_pri_config
)
summary(fit_gp_dense)
```

Let's examine the expected risk of each home under this model:

```{r}
pat |>
    mutate(pred_risk = fit_gp_dense$summary.fitted.values$mean) |>
    ggplot(aes(long, lat, col = as.factor(infestation), size = pred_risk)) +
    geom_point() +
    labs(x = NULL, y = NULL, col = 'Infestation status', size = "Pred. risk")
```

We find that the spatial model has identified that the lower part of the village has fewer infestations much better than
the model using only fixed effects.

### Further reading

Fitting spatial models using INLA may seem like a daunting task. However, if you are looking to harness the benefits of fully Bayesian inference
on a dataset containing more than a few hundred points, or need to fit many models (as is the case with adaptive sampling), its speed and relative 
ease of use is hard to beat. To learn more about spatial modeling with INLA, consider starting with these online, free resources:

- For a general overview using the R-INLA software, including some examples of spatial modeling, see Gómez-Rubio's online book 
(Bayesian inference with INLA)[https://becarioprecario.bitbucket.io/inla-gitbook/index.html]. This reference contains the information 
necessary to set prior distributions, sample from posterior predictive distributions, and handling missing values.
- For an intuitive overview of the INLA algorithm itself, see Kathryn Morrison's 
article (here)[https://www.precision-analytics.ca/articles/a-gentle-inla-tutorial/].
- For a deeper look into the SPDE approach but from a practical perspective, see the textbook 
(Advanced Spatial Modeling with Stochastic Partial Differential Equations Using R and INLA)[https://becarioprecario.bitbucket.io/spde-gitbook/].
This is probably the best reference to get started with applying INLA-SPDE to your own dataset.