---
title: Spatial modeling with INLA
subtitle: A short tutorial
author: B. K. M. Case
date: April 11, 2022
output: rmarkdown::github_document
---

Here we provide a short tutorial on fitting the types of spatial models used throughout the paper using the R-INLA package. 
INLA (Integrated Nested Laplacian Approximation) is an alternative to MCMC for fitting a wide variety of Bayesian models, 
and has become a popular option for geospatial models because of its speed and relative ease of use 
(see [Bakka et al. (2018)](https://wires.onlinelibrary.wiley.com/doi/pdf/10.1002/wics.1443) for a recent review).

### Load required packages and the data

```{r, include=FALSE}
knitr::opts_chunk$set(fig.path="examples")
```

```{r, message=FALSE, results=FALSE, warning=FALSE}
library(INLA)
library(tidyverse)
source("code/seq-sampling-helpers.R")

pat <- read_csv("anon-survey-data.csv") |>
    filter(village == "Paternito")
```

### Patterns of infestation

We are going to fit a spatial model to the infestation data in El Paternito. To begin exploring this data, we
can produce a plot of the binary infestation status of each home. Note that the latitude and longitude are not actual coordinates
here as they have been anonimized, but we can still treat them as $x$ and $y$ coordinates, respectively.

```{r}
ggplot(pat, aes(long, lat, col = as.factor(infestation))) +
    geom_point() +
    labs(x = '', y = '', col = 'Inf. Status')
```

We find that infested houses tend to be in middle and upper parts of the village, and often can be found
 close to other infested houses. 

### Traditional logistic regression

Before incorporating spatial effects, let's fit a model using only fixed effects from the covariate information.  
For simplicity, let's only use the covariates `bed_hygiene`, `num_dogs`, and `dist_perim`.
To fit such a model in R-INLA, we use the `inla` function:

```{r}
fx_only <- inla(infestation ~ bed_hygiene + num_dogs + dist_perim, data = pat, family = 'binomial')
summary(fx_only)
```

### Adding random effects in R-INLA

### Spatial smoothing and the SPDE approach

### Dense Gaussian Processes